Diffusive Homeostasis in a Self-Organizing Recurrent Neural Network \\ \begin{large} Spatially dependent Interaction as a Determinant of Neural Activity and Plasticity \end{large}

### 2017-03-02 draft.tex
.

Reducing oscillations meant to reduce height and frequency of the peak of the spectrum in the context of the presented power spectra. We managed to do so by increasing the time constant $\mathrm{\tau_{V_t}}$, as shown in Figure \ref{Peak_Prop_vs_Theta}. Moreover, by choosing $\mathrm{\sigma_{NO}}$, $\mathrm{\lambda}$ and $\mathrm{\alpha}$ in accordance to the fit shown in Figure \ref{Pow_Spec_Theta_vs_Analytic} (black curve), Equation \eqref{Pow_Spec_Theta_Final} provides a good prediction for the position and height of the peak. We smoothed the curve of the simulation data in plot (A) of Figure \ref{Peak_Prop_vs_Theta} by a window of $\mathrm{[-5,5]}$ data points for proper comparison, since the maximum of the power spectrum of simulation data is likely to overshoot the analytic prediction due to random fluctuations.  
\begin{figure}
%%% plot_osc_ampl_vs_tau_theta.py
\includegraphics[width=\textwidth]{../../plots/diff_hom/osc_ampl_vs_tau_th.png}
\caption{A: Maximum of the spectrum peak (see Figure \ref{Pow_Spec_Theta_vs_Analytic}). A smoothing window of $\mathrm{[-5,5]}$ data points around the maximum was used to reduce overestimation of the maximum due to random fluctuations. Blue line corresponds to Equation \eqref{Pow_Spec_Theta_Final}, fit values taken from black curve in Figure \ref{Pow_Spec_Theta_vs_Analytic}. B: Position of the peak on the frequency axis. Analytic prediction was acquired according to (A).}
\label{Peak_Prop_vs_Theta}
\end{figure} 
We eventually settled at a 1000-fold increased time constant to reduce oscillations as much as possible,  $\mathrm{\tau_{V_t}=2500\,s}$. Homeostasis still managed to keep excitatory network activity in the desired range, as shown in Figure \ref{full_sim_osci_slow_tau_th}, even given the slowness of threshold adaptation. No oscillating activity occurred despite a slight regularity of the time course of NO-concentration.  
\begin{figure}
%%% oscillating_activity_plot_full_time_only.py
\includegraphics[width=\textwidth]{../../plots/diff_hom/rate_NO_th_compl_large_tau.png}
\caption{A: Excitatory (blue) and and inhibitory (red) population rate, being controlled by diffusive homeostasis over the full simulation time. B,C: Mean NO-concentration over excitatory sites and mean excitatory threshold.}
\label{full_sim_osci_slow_tau_th}
\end{figure}
We took this setting as a basis for further analysis of network activity.

@@

\subsubsection{Spike-Train Statistics}
Balanced networks of inhibitory and excitatory populations of spiking LIF-neurons are known to show highly random, uncorrelated spiking behavior under sparse connectivity with a relatively high per-synapse connection strength \cite{Vreeswijk1996,Brunel2000}. This kind of asynchronous regular state is characterized by a population firing rate showing no regularities over time, though maintaining an overall stable level of activity. This regularity was shown to be achievable in the previous section after slowing down the rate of homeostatic adaption. Interspike intervals should be random on the level of single cells, approximately following the statistics of a Poisson process.   
\begin{figure}
%%% isi_plot.py
\includegraphics[width=\textwidth]{../../plots/Spike_Stats/isi_cv_compare.png}
\caption{A/C: Distribution of interspike intervals of excitatory/inhibitory neurons for diffusive and non-diffusive homeostasis, measured over $\mathrm{1000\,s}$. Only neurons with a mean firing rate in a certain range (see legend) were included into statistics to prevent overlap of different firing rates (except for non-diffusive homeostasis and excitatory neurons, all being very close to 3 Hz). Black lines are fits of exponential functions. B/D: Distribution of coefficients of variation. Each sample represents the CV of one excitatory/inhibitory neuron.}
\label{ISI_CV_compare}
\end{figure}
Figure \ref{ISI_CV_compare} shows the distribution of interspike intervals (ISIs) and the distribution of coefficients of variation ($\mathrm{CV\equiv\sigma/\mu}$) taken over the set of excitatory and inhibitory neurons, respectively. However, we restricted the statistics of ISIs to neurons whose mean firing rate fall into a window with a width of 1 Hz and a certain mean in order to provide comparability between the ISI distributions in the diffusive/non-diffusive case (see legend in Figure \ref{ISI_CV_compare}). An approximately exponential distribution of ISIs is apparently preserved for diffusive homeostasis (represented by straight lines in a log-plot), which is characteristic for Poisson processes \cite[p. 27]{Theor_Neur_Dayan}. For perfect poissonian spiking, the CV should give a value of 1. CVs that are smaller or larger, as seen in Figure \ref{ISI_CV_compare} B/D, indicate a tendency towards greater regularity or bunching, respectively. This observation is presumably due to the statistics of ISIs close to zero: Interestingly, while excitatory neurons have an  under-representation of short ISIs, the opposite is the case for inhibitory neurons. These types of behavior of LIF-models have been shown to depend on the CV of input the neuron receives: A smaller signal-to-noise ratio (larger CV) of neural input leads to a larger CV of ISIs for the output and an over-representation of small intervals \cite{Ostojic2011}. This relation fits nicely into our set of network parameters: Inhibitory neurons receive stronger recurrent inhibitory input ($\mathrm{-60\,mV}$ in total) compared to excitatory neurons ($\mathrm{-12\,mV}$). This lowers the signal-to-noise ratio for the inputs of inhibitory neurons, leading to the aforementioned statistical differences. Still, one can observe a dominant exponential component for both excitatory and inhibitory neurons.

We looked at potential correlations across neurons within both populations as an additional possibility of spike-train analysis. To do so, we calculated cross-correlations for all possible neuron pairings within a population, excluding auto-correlations. Figure \ref{Pop_Mean_Cross_Corr} shows the mean cross-correlation taken over all these pairs. Surprisingly, we found a significant increase in correlation across neurons for diffusive homeostasis. 
\begin{figure}
%plot_cross_corr.py
\includegraphics[width=\textwidth]{../../plots/Spike_Stats/pop_mean_cross_corr.png}
\caption{Mean of cross correlation over all pairs of neurons in the respective neuron group. Significantly higher correlation can be found in both groups for diffusive homeostasis. Spike data was taken from $\mathrm{1400-1500\,s}$ of the simulation.}
\label{Pop_Mean_Cross_Corr}
\end{figure}
One could consider two possibilities when speculating about the root of this discrepancy: Either, it is an immediate consequence of diffusive homeostasis, or it is indirectly related through other mechanisms present within the network. Regarding the first, it is certainly true that some amount of information about the activity of one neuron is transferred to another one in its neighborhood via diffusion. However, the response to fluctuations in the diffusive signal is on the order of seconds due to the time scale of the homeostatic adaption, while correlations in Figure \ref{Pop_Mean_Cross_Corr} are taking place within milliseconds. Thus, we concluded that a \emph{direct} link between homeostatic signaling and increased correlations can be regarded as implausible and that it must be an effect based on topological differences within the network. We report our investigation on these in Section \ref{Network_Topology_Section}.

@@

Despite of fluctuations within the firing rate on small time scales due to the inherent nature of---approximately---random spike generation, we wanted to know how strongly firing rates of single excitatory cells fluctuate over longer timescales of multiple seconds. This question is not only of relevance with respect to the ``stiffness" of homeostasis, but also relates to the plot in Figure \ref{Fir_Rate_Dist_Compare}: If firing rates fluctuate too much during the period over which means were calculated, the resulting distribution might be narrower than what one could expect from a distribution of momentary rates. A constant rate of an approximately randomly spiking neuron means that the dependence of fluctuations of rates onto the width of the averaging window should be identical to a homogeneous Poisson process of the same total mean firing rate. As bin sizes increase, possible long-term fluctuations of firing rates can then be identified as deviations from the variance one would expect from a homogeneous process. We tested this in Figure \ref{f_Var_vs_n_Bin}. (A) was taken from $\mathrm{1000-1500\,s}$, which is well inside the phase of diffusive homeostasis. As one can see, the statistics of the three representative neurons are well fitted by the predictions for homogeneous Poisson processes with the same mean rate, being a straight line given by $\mathrm{var = f_{\rm mean} n_{\rm bins}/T_{\rm total}}$, though slight deviations can be seen as a different factor of proportionality. (B) acts as an illustration showing a counterexample: Here, we included the entire time span of the simulation, including the transition between non-diffusive and diffusive homeostasis. This will cause a transition of firing rate between 3 Hz and a new rate different from 3 Hz for almost all of the neurons. Inhomogeneity is represented by the increasing positive relative deviation from the dashed line for small number of bins (i.e., large bin widths), even increasing again for a very few number of bins.
\begin{figure}
%%% plot_spt_var_vs_bin_width.py
\includegraphics[width=\textwidth]{../../plots/Spike_Stats/sim_f_var_vs_hom_poisson.png}
\caption{A: Relation between variance of the frequency signal (acquired by binning spikes) and the number of bins within the given time range of $\mathrm{1000-1500\,s}$, taken from three randomly picked excitatory neurons. B: Same procedure as in (A) for a randomly chosen excitatory neuron, but the time range was $\mathrm{0-1500\,s}$, which includes the initial phase of non-diffusive homeostasis. Dashed lines are the expected curve of a homogeneous Poisson process of the same mean rate in both plots.}
\label{f_Var_vs_n_Bin}
\end{figure}
We therefore concluded that diffusive homeostasis manages to retain firing rates on a constant level, not only population-wise but on a single-cell level as well. In particular, plastic modifications across the network due to STDP and synaptic growth/pruning---which were always acting for all the simulation results presented so far---appeared to be compensated for.

@@

A naturally emerging question when altering the diffusion constant is how the firing rate behaves in the absolute limit of infinitely fast diffusion. In fact, this case is quite easy to implement simulation-wise: One simply has to feed all NO-sources into a single scalar variable of NO concentration. This will provide the same NO readout for all excitatory neurons, which means that all excitatory thresholds change at the same rate all the time, only shifting their initial random distribution. Figure \ref{Fir_Rate_Dist_Instant_compare} shows the distribution for this special limiting case. The standard deviation for the excitatory population was $\mathrm{\sigma_{\rm inst.} = 1.79\, Hz}$ and the skewness $\mathrm{v_{\rm inst.} = 1.51}$ ($\mathrm{v_{\rm inst.,log} = -0.52}$), which makes the asymmetry slightly more pronounced than in Figure \ref{Fir_Rate_Dist_Compare}.  
\begin{figure}
%%% frequ_dist_instant_diff.py
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/instant_diff_fir_dist_e.png}
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/instant_diff_fir_dist_i.png}
\caption{Distribution of firing rates for instantaneous diffusion for excitatory (A,B) and inhibitory (A*,B*) population. Data was taken from 10 simulations and $\mathrm{1000-1500\,s}$.}
\label{Fir_Rate_Dist_Instant_compare}
\end{figure}

@@

\subsection{Network Topology}\label{Network_Topology_Section}
We have only presented our analysis of network activity so far. In this section we present results that relate to features of synaptic topology. Since the SORN we used for all simulations only included plastic mechanisms affecting recurrent excitatory connections, these were the subject of our investigations. Moreover, in contrast to the previous section, we did not explicitly aim for the emergence of a \emph{new} feature of synaptic topology, but rather sought to recover properties that had been found in earlier versions of the LIF-SORN.

\subsubsection{Excitatory Connection Fraction}
In the previous version of the LIF-SORN, initializing the network with zero recurrent excitatory connections led to a monotonically increasing but saturating excitatory-to-excitatory connection fraction (CF). The growth rate was tuned such that the terminal CF settled at $\mathrm{10\%}$, see \cite[p. 8]{SORN_Paper}. Omitting the distance dependence of connection probability for synaptic growth led to a slightly higher connection fraction. We ran a full simulation under diffusive homeostasis to compare these previous results to the diffusive case, including the growth phase by setting the target concentration to an appropriate value that was determined in a previous simulation with non-diffusive homeostasis. The resulting time course of CF is depicted in Figure \ref{CF_plot}.
%%% plot_CF.py
\includegraphics[width=\textwidth]{../../plots/syn_topology/CF_plot.png}
\caption{Connection fractions of recurrent excitatory synapses. Simulation protocols include all four possible combinations of diffusive/non-diffusive homeostasis and distance-dependent/non-distance-dependent synaptic growth, each variant retained for the entire simulation time. Data from 10 trials, linewidth represents standard error.}
\label{CF_plot}
Diffusive homeostasis caused the CF to slightly overshoot its final fraction in the growth phase. To understand this kind of behavior, we recalled the two mechanisms that directly determine changes within the connection fraction, namely synaptic growth and pruning. The average change of connection fraction is simply proportional to the difference between newly created synapses and those that were removed. Though fluctuating, the synaptic growth rate was kept at a constant level of 920 synapses/second. Any differences in Figure \ref{CF_plot} must therefore originate in different pruning rates. Figure \ref{STDP_change} shows the mean and standard deviation of weight changes due to STDP between two normalization steps. Non-diffusive homeostasis has a higher mean as well as a broader distribution of weight fluctuations throughout the beginning of the simulation. However, the difference of mean changes is orders of magnitude smaller than the width of fluctuations. Thus, while it might seem counter-intuitive that a higher positive mean weight change comes with a smaller chance of synaptic survival, we attribute the main cause of overshooting in Figure \ref{CF_plot} to the increased fluctuation of weights since it raises the chance of going below the pruning threshold within a given time interval.
%%% stdp_change_time.py
\includegraphics[width=\textwidth]{../../plots/syn_topology/av_stdp_change_time.png}
\caption{Top: Mean change of synaptic weights (growth and pruning excluded) between two normalization steps. Bottom: Standard deviation for the same data. Both curves correspond to those in Figure \ref{CF_plot} with the same color.}
\label{STDP_change}
Apart from reproducing the desired connection fraction, previous versions of the LIF-SORN showed an over-representation of bidirectional connections compared to a random graph with equal connectivity \cite{SORN_Paper}. This experimentally observed feature required the presence of a distance dependent connection probability, as described in section \ref{network simulation}. The absence of this breaking of spatial homogeneity even led to an under-representation of bidirectional connections, which is known to be an effect of STDP in recurrent networks \cite{Syn_Plast_Abbott}. The separation between simulations with and without a spatial connection profile is retained, as shown in Figure \ref{RC_CFB_plot}. 
%%% plot_RC_CFB.py
\includegraphics[width=\textwidth]{../../plots/syn_topology/RC_CFB_plot.png}
\caption{Fraction of bidirectional connections normalized over the expected fraction for a random graph with equal total connection fraction. Data from 10 trials, linewidth represents standard error.}
\label{RC_CFB_plot}
\subsubsection{Distribution of Synaptic Weights} \label{Syn_Weight_Dist_Section}
A well-studied property of neural networks is the distribution of synaptic weights. Experimental studies in hippocampus and cortical regions suggest heavy-tailed log-normal-like distribution of synaptic efficacies \cite{Song_Connectivity_2005,Lisman_Synapses_1993,Yasumatsu_Synapses_2008,Loewenstein_Spine_Sizes}. Theoretical models are mainly based on a combination of multiplicative and additive weight dynamics \cite{Loewenstein_Spine_Sizes,Statman_Synapses_2014}, which is in line with our implementation of multiplicative normalization and additive STDP. Figure \ref{Weight_Dist} compares the resulting distributions. We found that we can retain a log-normal-like distribution in combination with diffusive homeostasis. A slightly larger variance can be observed. A possible explanation refers back to our observations in Figure \ref{STDP_change}: Regarding STDP as a random process, a broader distribution of weights can be explained by a smaller potentiating effect and a decreased variance of STDP-fluctuations, which is both the case for diffusive homeostasis in the beginning of the simulation. Still, it seems unlikely that a possible effect of these differences is still present at $\mathrm{t=1500\,s}$.
%%% plot_weight_dist.py
\includegraphics[width=\textwidth]{../../plots/syn_topology/syn_weight_dist.png}
\caption{Distribution of decadic logarithm of excitatory synaptic weights at $\mathrm{t=1500\,s}$ for diffusive and non-diffusive homeostasis (both with distance-dependent connectivity). Gaussian fits returned $\mathrm{\mu_{\rm diff.}=-0.17\, \mathrm{log}_{10}(mV)}$, $\mathrm{\mu_{\rm non-diff.}=-0.087\, \mathrm{log}_{10}(mV)}$, $\mathrm{\sigma_{\rm diff.}=0.64\, log_{10}(mV)}$ and $\mathrm{\sigma_{non-diff.}=0.42\, log_{10}(mV)}$.}
\label{Weight_Dist}

>>>>

Reducing oscillations meant to reduce height and frequency of the peak of the spectrum in the context of the presented power spectra. We managed to do so by increasing the time constant $\mathrm{\tau_{V_t}}$, as shown in Figure \ref{Peak_Prop_vs_Theta}. Moreover, by choosing $\mathrm{\sigma_{NO}}$, $\mathrm{\lambda}$ and $\mathrm{\alpha}$ in accordance to the fit shown in Figure \ref{Pow_Spec_Theta_vs_Analytic} (black curve), Equation \eqref{Pow_Spec_Theta_Final} provides a good prediction for the position and height of the peak. We smoothed the curve of the simulation data in plot (A) of Figure \ref{Peak_Prop_vs_Theta} by a window of $\mathrm{[-5,5]}$ data points for proper comparison, since the maximum of the power spectrum of simulation data is likely to overshoot the analytic prediction due to random fluctuations.  
\begin{figure}
%%% plot_osc_ampl_vs_tau_theta.py
\includegraphics[width=\textwidth]{../../plots/diff_hom/osc_ampl_vs_tau_th.png}
\caption{A: Maximum of the spectrum peak (see Figure \ref{Pow_Spec_Theta_vs_Analytic}). A smoothing window of $\mathrm{[-5,5]}$ data points around the maximum was used to reduce overestimation of the maximum due to random fluctuations. Fit values for analytic prediction were taken from black curve in Figure \ref{Pow_Spec_Theta_vs_Analytic}. B: Position of the peak on the frequency axis.}
\label{Peak_Prop_vs_Theta}
\end{figure} 
We eventually settled at a $\rm 1000$-fold increased time constant to reduce oscillations as much as possible,  $\mathrm{\tau_{V_t}=2500\,s}$. Homeostasis still managed to keep excitatory network activity in the desired range, as shown in Figure \ref{full_sim_osci_slow_tau_th}, even given the slowness of threshold adaptation. No oscillating activity occurred despite a slight regularity of the time course of NO-concentration.  
\begin{figure}
%%% oscillating_activity_plot_full_time_only.py
\includegraphics[width=\textwidth]{../../plots/diff_hom/rate_NO_th_compl_large_tau.png}
\caption{A: Excitatory (blue) and and inhibitory (red) population rate, being controlled by diffusive homeostasis over the full simulation time. B,C: Mean NO-concentration over excitatory sites and mean excitatory threshold.}
\label{full_sim_osci_slow_tau_th}
\end{figure}
We took this setting as a basis for further analysis of network activity.

@@

\subsubsection{Spike-Train Statistics}
Balanced networks of inhibitory and excitatory populations of spiking LIF-neurons are known to show highly random, uncorrelated spiking behavior under sparse connectivity with a relatively high per-synapse connection strength \cite{Vreeswijk1996,Brunel2000}. This kind of asynchronous irregular state is characterized by a population firing rate showing no regularities over time, though maintaining an overall stable level of activity. This regularity was shown to be achievable in the previous section after slowing down the rate of homeostatic adaption. Interspike intervals should be random on the level of single cells, approximately following the statistics of a Poisson process.   
\begin{figure}
%%% isi_plot.py
\includegraphics[width=\textwidth]{../../plots/Spike_Stats/isi_cv_compare.png}
\caption{A/C: Distribution of interspike intervals of excitatory/inhibitory neurons for diffusive and non-diffusive homeostasis, measured over $\mathrm{1000\,s}$. Only neurons with a mean firing rate in a certain range (see legend) were included into statistics to prevent overlap of different firing rates (except for non-diffusive homeostasis and excitatory neurons, all being very close to 3 Hz). B/D: Distribution of coefficients of variation. Each sample represents the CV of one excitatory/inhibitory neuron.}
\label{ISI_CV_compare}
\end{figure}
Figure \ref{ISI_CV_compare} shows the distribution of interspike intervals (ISIs) and the distribution of coefficients of variation ($\mathrm{CV\equiv\sigma/\mu}$) taken over the set of excitatory and inhibitory neurons. However, we restricted the statistics of ISIs to neurons whose mean firing rate fall into a window with a width of 1 Hz and a certain mean in order to provide comparability between the ISI distributions in the diffusive/non-diffusive case (see legend in Figure \ref{ISI_CV_compare}). An approximately exponential distribution of ISIs is apparently preserved for diffusive homeostasis (represented by straight lines in a log-plot), which is characteristic for Poisson processes \cite[p. 27]{Theor_Neur_Dayan}. For perfect poissonian spiking, the CV should give a value of 1. CVs that are smaller or larger, as seen in Figure \ref{ISI_CV_compare} B/D, indicate a tendency towards greater regularity or bunching, respectively. This observation is presumably due to the statistics of ISIs close to zero: Interestingly, while excitatory neurons have an  under-representation of short ISIs, the opposite is the case for inhibitory neurons. These types of behavior of LIF-models have been shown to depend on the amount of fluctuations of the input the neuron receives: A larger CV of neural input leads to a larger CV of ISIs for the output and an over-representation of small intervals \cite{Ostojic2011}. By the CV of the input we mean the ratio between standard deviation and mean of postsynaptic current due to presynaptic spiking events (under the assumption of a stationary mean and variance). This relation between fluctuating input and spiking statistics fits nicely into our set of network parameters: Inhibitory neurons receive stronger recurrent inhibitory input ($\mathrm{-60\,mV}$ in total) compared to excitatory neurons ($\mathrm{-12\,mV}$). This lowers the signal-to-noise ratio for the inputs of inhibitory neurons, leading to the aforementioned statistical differences. Still, one can observe a dominant exponential component for both excitatory and inhibitory neurons.

We also considered potential correlations across neurons within both populations as an additional possibility of spike-train analysis. To do so, we calculated cross-correlations for all possible neuron pairings within a population, excluding auto-correlations. Figure \ref{Pop_Mean_Cross_Corr} shows the mean cross-correlation taken over all these pairs. Surprisingly, we found a significant increase in correlation across neurons for diffusive homeostasis. 
\begin{figure}
%plot_cross_corr.py
\includegraphics[width=\textwidth]{../../plots/Spike_Stats/pop_mean_cross_corr.png}
\caption{Mean of cross correlation over all pairs of neurons in the respective neuron group. Significantly higher correlation can be found in both groups for diffusive homeostasis. Spike data was taken from $\mathrm{1400-1500\,s}$ of the simulation.}
\label{Pop_Mean_Cross_Corr}
\end{figure}
One could consider two possibilities when speculating about the root of this discrepancy: Either, it is an immediate consequence of diffusive homeostasis, or it is indirectly related through other mechanisms present within the network. Regarding the first, it is certainly true that some amount of information about the activity of one neuron is transferred to another one in its neighborhood via diffusion. However, the response to fluctuations in the diffusive signal is on the order of seconds due to the time scale of the homeostatic adaption, while correlations in Figure \ref{Pop_Mean_Cross_Corr} are taking place within milliseconds. Thus, we concluded that a \emph{direct} link between homeostatic signaling and increased correlations can be regarded as implausible and that it must be an effect based on topological differences within the network. We report our investigation on these in Section \ref{Network_Topology_Section}.

@@

Despite of fluctuations within the firing rate on small time scales due to the inherent nature of---approximately---random spike generation, we wanted to know how strongly firing rates of single excitatory cells fluctuate over longer timescales of multiple seconds. This question is not only of relevance with respect to the ``stiffness" of homeostasis, but also relates to the plot in Figure \ref{Fir_Rate_Dist_Compare}: If firing rates fluctuate too much during the period over which means were calculated, the resulting distribution might be narrower than what one could expect from a distribution of momentary rates. A constant rate of an approximately randomly spiking neuron means that the dependence of fluctuations of rates on the width of the averaging window should be identical to a homogeneous Poisson process of the same total mean firing rate. As bin sizes increase, possible long-term fluctuations of firing rates can then be identified as deviations from the variance one would expect from a homogeneous process. We tested this in Figure \ref{f_Var_vs_n_Bin}. (A) was taken from $\mathrm{1000-1500\,s}$, which is well inside the phase of diffusive homeostasis. As one can see, the statistics of the three representative neurons are well fitted by the predictions for homogeneous Poisson processes with the same mean rate, being a straight line given by $\mathrm{var = f_{\rm mean} n_{\rm bins}/T_{\rm total}}$, though slight deviations can be seen as a different factor of proportionality. (B) acts as an illustration showing a counterexample: Here, we included the entire time span of the simulation, including the transition between non-diffusive and diffusive homeostasis. This will cause a transition of firing rate between 3 Hz and a new rate different from 3 Hz for almost all of the neurons. Inhomogeneity is represented by the increasing positive relative deviation from the dashed line for small number of bins (i.e., large bin widths), even increasing again for a very few number of bins.
\begin{figure}
%%% plot_spt_var_vs_bin_width.py
\includegraphics[width=\textwidth]{../../plots/Spike_Stats/sim_f_var_vs_hom_poisson.png}
\caption{A: Relation between variance of the frequency signal (acquired by binning spikes) and the number of bins within the given time range of $\mathrm{1000-1500\,s}$, taken from three randomly picked excitatory neurons. B: Same procedure as in (A) for a randomly chosen excitatory neuron, but the time range was $\mathrm{0-1500\,s}$, which includes the initial phase of non-diffusive homeostasis. Dashed lines are the expected curve of a homogeneous Poisson process of the same mean rate in both plots.}
\label{f_Var_vs_n_Bin}
\end{figure}
We therefore concluded that diffusive homeostasis manages to retain firing rates on a constant level, not only population-wise but on a single-cell level as well. In particular, plastic modifications across the network due to STDP and synaptic growth/pruning---which were always acting for all the simulation results presented so far---appeared to be compensated for.

@@

A naturally emerging question when altering the diffusion constant is how the firing rate behaves in the absolute limit of infinitely fast diffusion. In fact, this case is quite easy to implement simulation-wise: One simply has to feed all NO-sources into a single scalar variable of NO concentration. This will provide the same NO readout for all excitatory neurons, which means that all excitatory thresholds change at the same rate all the time, only shifting their initial random distribution. Figure \ref{Fir_Rate_Dist_Instant_compare} shows the distribution for this special limiting case. The standard deviation for the excitatory population was $\mathrm{\sigma_{\rm inst.} = 1.79\, Hz}$ and the skewness $\mathrm{v_{\rm inst.} = 1.51}$ ($\mathrm{v_{\rm inst.,log} = -0.52}$), which makes the asymmetry slightly more pronounced than in Figure \ref{Fir_Rate_Dist_Compare}.  
\begin{figure}
%%% frequ_dist_instant_diff.py
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/instant_diff_fir_dist_e.png}
\includegraphics[width=\textwidth]{../../plots/firing_rate_dists/instant_diff_fir_dist_i.png}
\caption{Distribution of firing rates for instantaneous diffusion for excitatory (A,B) and inhibitory (A*,B*) population. Data was taken from 10 simulations and $\mathrm{1000\,s \leq t \leq 1500\,s}$.}
\label{Fir_Rate_Dist_Instant_compare}
\end{figure}

@@

\subsection{Network Topology}\label{Network_Topology_Section}
We have only presented our analysis of network activity so far. In this section we present results that relate to features of network topology. Since the SORN we used for all simulations only included plastic mechanisms affecting recurrent excitatory connections, these were the subject of our investigations. Moreover, in contrast to the previous section, we did not explicitly aim for the emergence of a \emph{new} feature of network topology, but rather sought to recover properties that had been found in earlier versions of the LIF-SORN.

\subsubsection{Excitatory Connection Fraction}
In the previous version of the LIF-SORN, initializing the network with zero recurrent excitatory connections led to a overall increasing but saturating time course of the excitatory-to-excitatory connection fraction (CF). The growth rate was tuned such that the terminal CF settled at $\mathrm{10\%}$, see \cite[p. 8]{SORN_Paper}. Omitting the distance dependence of connection probability for synaptic growth led to a slightly higher connection fraction. We ran a full simulation under diffusive homeostasis to compare these previous results to the diffusive case, including the growth phase by setting the target concentration to an appropriate value that was determined in a previous simulation with non-diffusive homeostasis. The resulting time course of CF is depicted in Figure \ref{CF_plot}.
\begin{figure}
%%% plot_CF.py
\includegraphics[width=\textwidth]{../../plots/syn_topology/CF_plot.png}
\caption{Connection fractions of recurrent excitatory synapses. Simulation protocols include all four possible combinations of diffusive/non-diffusive homeostasis and distance-dependent/non-distance-dependent synaptic growth, each variant retained for the entire simulation time. Data from 10 trials, linewidth represents standard error.}
\label{CF_plot}
\end{figure}
Diffusive homeostasis caused the CF to slightly overshoot its final fraction in the growth phase. To understand this kind of behavior, we recalled the two mechanisms that directly determine changes within the connection fraction, namely synaptic creation and pruning. The average change of connection fraction is simply proportional to the difference between newly created synapses and those that were removed. Though fluctuating, the synaptic growth rate was kept at a constant level of 920 synapses/second. Any differences in Figure \ref{CF_plot} must therefore originate in different pruning rates. Figure \ref{STDP_change} shows the mean and standard deviation of weight changes due to STDP between two normalization steps. Non-diffusive homeostasis has a higher mean as well as a broader distribution of weight fluctuations throughout the beginning of the simulation. However, the difference of mean changes is orders of magnitude smaller than the width of fluctuations. Thus, while it might seem counter-intuitive that a higher positive mean weight change comes with a smaller chance of synaptic survival, we attribute the main cause of overshooting in Figure \ref{CF_plot} to the increased fluctuation of weights since it raises the chance of going below the pruning threshold within a given time interval.
\begin{figure}
%%% stdp_change_time.py
\includegraphics[width=\textwidth]{../../plots/syn_topology/av_stdp_change_time.png}
\caption{Top: Mean change of synaptic weights (creation and pruning excluded) between two normalization steps. Bottom: Standard deviation for the same data. Both curves correspond to those in Figure \ref{CF_plot} with the same color.}
\label{STDP_change}
\end{figure}

Apart from reproducing the desired connection fraction, previous versions of the LIF-SORN showed an over-representation of bidirectional connections compared to a random graph with equal connectivity \cite{SORN_Paper}. This experimentally observed feature required the presence of a distance dependent connection probability, as described in section \ref{network simulation}. The theoretical background of this effect was laid out by Miner and Triesch \cite{SORN_Paper} and further generalized by Hoffmann and Triesch \cite{Hoffmann_2017}. The result of these investigations was that any deviation from a perfectly uniform connection probability that preserves symmetry of these connection probabilities will necessarily lead to an over-representation of bidirectional connections. We shall explain this effect in more detail.

First, assume that the connection probability $p_{\rm hom.}$ between every pair of neurons is equal. The resulting structure is called an Erd\H{o}s-Renyi graph and the bidirectional connection probability is simply $p_{\rm hom.}^2$. Now, we consider the case where each neuron $i$ connects to another neuron $j$ with a certain probability $p_{ij}$ and where neuron $j$ connects with equal probability $p_{ji}=p_{ij}$ to neuron $i$. We further assume that all these connection probabilities are drawn from the same distribution. The bidirectional connection probability between a random pair of neurons is thus
\begin{equation}
p_{\rm bidir.} = \mathrm{E[}p_{ij}p_{ji}\mathrm{]} = \mathrm{E[}p_{ij}^2\mathrm{]}
\label{bidir_prob}
\end{equation}
where $\rm E[\cdot ]$ denotes the expected value. The \textit{relative} bidirectional connectivity now compares this value to the expected bidirectional connection probability of an Erd\H{o}s-Renyi graph with the same unidirectional overall connection probability. In such a graph, the unidirectional connection probability is $\mathrm{E[}p_{ij}\mathrm{]}$ and thus, the bidirectional connection is
\begin{equation}
p_{\rm bidir. rand.} = \mathrm{E[}p_{ij}\mathrm{]}^2 \; .
\label{bidir_rand_prob}
\end{equation}
The ration we would like to calculate thus is
\begin{equation}
r_{CFB} = \frac{\mathrm{E[}p_{ij}^2\mathrm{]}}{\mathrm{E[}p_{ij}\mathrm{]}^2} \; .
\end{equation}
Using Jensen's inequality and the fact that the squaring function is convex, Hoffmann and Triesch showed that for any distribution of connection probabilities with a non-vanishing variance $\mathrm{Var[}p_{ij}\mathrm{]} > 0$, $r_{CFB}$ will be greater or equal to one. This provided an explanation why a non-uniform, distance-dependent connection profile with reciprocally symmetric connection probabilities leads to an over-representation of bidirectional connections.

In the plastic network, the absence of this breaking of spatial homogeneity even led to an under-representation of bidirectional connections, which is known to be an effect of STDP in recurrent networks \cite{Syn_Plast_Abbott}. The separation between simulations with and without a spatial connection profile is retained after having implemented diffusive homeostasis, as shown in Figure \ref{RC_CFB_plot}. 
\begin{figure}
%%% plot_RC_CFB.py
\includegraphics[width=\textwidth]{../../plots/syn_topology/RC_CFB_plot.png}
\caption{Fraction of bidirectional connections normalized over the expected fraction for a random graph with equal total connection fraction. Data from 10 trials, linewidth represents standard error.}
\label{RC_CFB_plot}
\end{figure}
\subsubsection{Distribution of Synaptic Weights} \label{Syn_Weight_Dist_Section}
A well-studied property of neural networks is the distribution of synaptic weights. Experimental studies in hippocampus and cortical regions suggest heavy-tailed log-normal-like distributions of excitatory synaptic efficacies \cite{Song_Connectivity_2005,Lisman_Synapses_1993,Yasumatsu_Synapses_2008,Loewenstein_Spine_Sizes}. Theoretical models are mainly based on a combination of multiplicative and additive weight dynamics \cite{Loewenstein_Spine_Sizes,Statman_Synapses_2014}, which is in line with our implementation of multiplicative normalization and additive STDP. Figure \ref{Weight_Dist} compares the resulting distributions. We found that we can retain a log-normal-like distribution in combination with diffusive homeostasis. A slightly larger variance can be observed. A possible explanation refers back to our observations in Figure \ref{STDP_change}: Regarding STDP as a combination of a multiplicative and an additive random process, a broader distribution of weights can be explained by a smaller potentiating effect, which is the case for diffusive homeostasis in the beginning of the simulation. Still, it seems unlikely that a possible effect of these differences is still present at $\mathrm{t=1500\,s}$.
\begin{figure}
%%% plot_weight_dist.py
\includegraphics[width=\textwidth]{../../plots/syn_topology/syn_weight_dist.png}
\caption{Distribution of decadic logarithm of excitatory synaptic weights at $\mathrm{t=1500\,s}$ for diffusive and non-diffusive homeostasis (both with distance-dependent connectivity). Gaussian fits returned $\mathrm{\mu_{\rm diff.}=-0.17\, \mathrm{log}_{10}(mV)}$, $\mathrm{\mu_{\rm non-diff.}=-0.087\, \mathrm{log}_{10}(mV)}$, $\mathrm{\sigma_{\rm diff.}=0.64\, log_{10}(mV)}$ and $\mathrm{\sigma_{non-diff.}=0.42\, log_{10}(mV)}$.}
\label{Weight_Dist}
\end{figure}
